\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-26 paper.tex

\section{Example inferred ARGs}
\label{sec-example-inferred-args}

\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} Inference of sample-resolved ARGs for 11
\textit{Drosophila melanogaster} DNA sequences over a 2.4kb
region of the ADH locus~\citep{kreitman1983nucleotide}.
Results for four different methods based on
% FIXME: this is the wrong place to be summarising. Name the methods
% here and summarise in the text
(A) parsimony, (B) sampling the SMC$^\prime$, (C) copying from
inferred ancestral haplotypes, and (D) correlated local tree estimation.
Edge colours are based on the time of the edge's child node
(lighter: older; darker: younger).
X and Y positions of graph nodes are arbitrary;
line width and node colour are as described in Fig.~\ref{fig-simplification}.
Graphics on the bottom row show the genome positions, relative to the start of the ADH gene,
for each graph edge from the corresponding ARG. Edge intervals are drawn as horizontal lines,
stacked in time order (edges with youngest children at the bottom);
vertical dashed lines indicate breakpoints between local trees.
}
\end{figure}

The scalability gains made by recent ARG inference methods have been, in part, due to inferring
lower levels of precision about recombination. Indeed, a common statement in the literature is
that modern tools such as \tsinfer~\citep{kelleher2019inferring} and \relate~\citep{speidel2019method}
construct not an ARG but ``only'' a sequence of local
trees \citep[e.g.][]{hejase2020summary, brandt2021evaluation, ignatieva2021kwarg}. While this statement
is correct inasmuch as neither method attempts to infer the explicit recombination events that cause
adjacent trees to differ, it is certainly not the case that they infer independent local trees.
In fact, both methods take advantage of genome-wide mutation information to construct genealogies,
and as a result nodes and edges persist across multiple local trees.
Their results can therefore be thought of as an inheritance graph, and represented in the gARG format.

To illustrate the varying levels of information that can be captured in a gARG,
Fig.~\ref{fig-inferred-args} presents graphical representations of genealogies output by two tools that
\emph{do} explicitly infer recombination events, namely \kwarg~\citep{ignatieva2021kwarg} and
\argweaver~\citep{rasmussen2014genome},
and two which do not (\tsinfer\ and \relate).
All inferences are based on the
same underlying dataset from \citet{kreitman1983nucleotide}, a standard benchmark in the classical
ARG literature, for which the minimum number of recombination breakpoints in the absence of recurrent
mutation is 7~\citep{song2003parsimonious}. This dataset comprises 43 biallelic SNPs spanning 2.4Kb of the
\emph{D.~melanogaster} ADH locus on chromosome 2L. Where required for inference purposes we
assume a mutation rate of
$5.49\times10^{-9}$ per site per generation
and a recombination rate of
$8.4\times10^{-9}$ per site per generation
\citep{adrion2020community,lauterbur2022expanding},
together with
a constant effective population size of 1,000,000.
To focus on the ARG topology rather than estimates of node times, the Y position of each node
is arbitrary, taken from the ranked order of nodes output by each software method.
The version numbers of inference software, the specific inference parameters
needed to reproduce the ARGs, and (where necessary) functions to convert outputs to the
\texttt{tskit} gARG format, are available from our GitHub repository.

Note that both \kwarg\ and \argweaver\ produce many ARGs
(those shown in Fig.~\ref{fig-inferred-args}A,B are arbitrarily selected examples).
This stems from the fact that \kwarg\ searches the space of plausible ARGs,
outputting minimal ones using heuristics, and \argweaver\ samples
from the posterior distribution over
ARGs given the data and a fixed model (a discretised version of the SMC$^\prime$ \citep{marjoram2006fast},
an approximation of the full coalescent with recombination).
In contrast, \tsinfer\ and \relate\ (Fig.~\ref{fig-inferred-args}C,D) are
deterministic methods, aiming to find a single ``best guess'' ARG.
The ability of a gARG to encode the effects of recombination, rather
than having to precisely specify each event, is an important contributor to this aim,
as most recombination events cannot be uniquely identified from the data.
Indeed, there is considerable variation in the number of recombinations inferred by
different methods, with e.g.~\argweaver\ (Fig.~\ref{fig-inferred-args}B) suggesting
far more than are required under minimal parsimony assumptions (Fig.~\ref{fig-inferred-args}A).

% Here we describe recombination features
A rough feel for the amount of recombination in each ARG is given by the distribution
of number of parents per node, as highlighted by the node colour scheme.
In Fig.~\ref{fig-inferred-args}A,B, each recombination event is tagged by
a node with exactly 2 parents and one child. Assuming that the breakpoint for each event
% TODO "infinite breakpoints" is goign to confuse people. What are we
% clarifying here? Citation necessary if we keep.
occurs at a different location (the ``infinite breakpoints'' model), then the
number of breakpoints should exactly equal the number of 2-parent (brown) nodes in the plot.
In contrast, Figure Fig.~\ref{fig-inferred-args}C,D have ARG nodes with multiple
parents and multiple children. These indicate deliberate imprecision, where one
or more recombinations have been ``pushed down'' onto a more recent, identifiable node.
Several of these nodes have more than two parents (in red), representing
multiple recombinations inferred to have occurred in an unresolvable order.
Nevertheless, the number of breakpoints in these two ARGs is smaller than the
number of multiple-parent ARG nodes, showing that several multiple-parent nodes
must share breakpoint positions. The sharing of breakpoint locations
between these nodes indicates that in \tsinfer\ and \relate\ ARGs it may be
difficult to condense each transition between trees into a set of simple SPR moves.
Overall, Fig.~\ref{fig-inferred-args} shows that \tsinfer\ and \relate\ ARGs contain a level
of detail about recombination that lies somewhere between a sequence of unrelated local
trees on one extreme and an ARG with precisely specified recombination events on the other.

% Here we move on to coalescence
As for inferred coalescences, the shading (``coalescent span'')
reveals that many non-recombinant (green) nodes are
locally unary (see Section \ref{sec-locally-unary-edges}). In other words,
in some regions of the genome these nodes are present in the ancestry without
corresponding to coalescence of ancestral material. As argued previously, this can provide useful
information about shared ancestry. Note however, that such unary segments are not inferred
in the \relate\ ARG: other than the samples it has entirely solid nodes.
Another clear difference involves coalescent nodes with more than two children (polytomies);
in Appendix \ref{sec-cell-lineages-and-args} we argue that this reflects
uncertainty in the order of coalescent events, in an equivalent
manner to the imprecision associated with recombination nodes having more
than two parents. Only \tsinfer\ creates such polytomies, the clearest of which
involves the nodes \noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F}. These
happen to have identical sequences, but because \kwarg, \argweaver, and \relate\ require
bifurcating trees, they are forced to pick an arbitrary order of branching
(hence Fig.~\ref{fig-inferred-args}A,B disagrees with Fig.~\ref{fig-inferred-args}D
in this order).
More generally, the number of nodes in Fig.~\ref{fig-inferred-args}C
that have more than two children hints at considerable uncertainty in
the order of branching elsewhere in the ARG.

%% possible TODO - mention something about multiple local roots in
% tsinfer and Relate versus the other two.

%The example of successive simplification of a simulated ARG
%in Fig.~\ref{fig-simplification} illustrates the point
%that is is possible to \emph{store} arbitrary levels of precision
%about recombination in a gARG.
%It is also possible to \emph{infer} various levels of precision about
%recombination.
%To illustrate this, and to demonstrate the structural
%heterogeneity of the ARGs inferred  by current methods,
%Fig.~\ref{fig-inferred-args} shows the results of ARG inference
% To illustrate these points and to emphasise the heterogeneity
% of the ARGs inferred by current methods, we inferred ARGs for the
%for the classical \citet{kreitman1983nucleotide} benchmark dataset
%using  \kwarg~\citep{ignatieva2021kwarg},
%\argweaver~\citep{rasmussen2014genome,hubisz2020inference},
%\tsinfer~\citep{kelleher2019inferring},
%and \relate~\citep{speidel2019method}.
% While there is some consistency among the inferred ARGs (such as the samples
% FR-F, WA-F and Af-F being grouped into the same clade), there
% are also considerable differences.
%The minimum number
%of recombination breakpoints required in the absence of recurrent mutation
%is 7 for this dataset~\citep{song2003parsimonious},
%and the number of inferred breakpoints
%under the input mutation and recombination rates
%varies from 2 (\relate) to 17 (\argweaver).
% TODO this needs a bit of rewriting with the rest of the paper in mind
% kwarg and argweaver are inferring eARGs, but we need to be careful
% not to confuse people.
%The ARGs inferred by the different methods also have major differences
%in how recombination information is represented.
%Essentially,
%\kwarg\ and \argweaver\ infer eARGs that contain
%specific recombination \emph{events} for each breakpoint.
%On the other hand, \tsinfer\ and
%\relate\ do not infer events but rather
%infer gARGs with by differing levels of precision about recombination.

% TODO add additional explanation of the bottom row, if enough isn't written below
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/386
The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
the genome to which graph edges are shared between multiple trees.
This illustrates that both \tsinfer\ and (to a lesser extent) \relate\ have edges
that span multiple tree boundaries, indicating that they are not inferring
a series of \emph{unrelated} local trees.
%(i.e., with either unlabelled internal
%nodes or differently labelled nodes in each tree).
For example, both infer that \noderef{Af-f}, \noderef{Fr-f}, and \noderef{Wa-f}
share a common ancestor along
the entire sequence. However, in comparison to \kwarg\ and \argweaver\ neither
method results in extensive node sharing in the
oldest time periods. This primarily reflects the fact that they do not encode topology
above the local MRCAs, also indicated by multiple roots in the ARG
(nodes with zero parents, in blue), compared to the
single ARG root in the \kwarg\ and \argweaver\ genealogies. In terms of inference,
\tsinfer\ and \relate\ show a pragmatic representation: we do not expect
genetic variation to be very informative about topological relationships
between nodes that that are older than the local MRCAs.

>>>>

\section{Example inferred ARGs}
\label{sec-example-inferred-args}
The scalability gains made by recent ARG inference methods
such as \relate~\citep{speidel2019method} and
\tsinfer~\citep{kelleher2019inferring} have
have been, in part, due to inferring lower levels of precision about
recombination than classical methods.
Neither method infers
explicit recombination events, and therefore their outputs cannot
be described using the classical eARG formalisms (Section~\ref{sec-eARG}).
% Could do, something like
% this mismatch has led some to characterise their output as a
% series of unrelated local trees [cite], which lacks nuance of the reality.
% Or maybe we don't bother: what's to be gained from pointing out people's
% mistakes here? Hopefully the rich detail of reality will come through
% and replace old ideas.
Nonetheless, both methods produce estimates in which nodes and edges
persist across multiple trees, naturally captured by the gARG approach.
To illustrate the varying levels of information captured
by current methods, and some qualitative differences between them,
Fig.~\ref{fig-inferred-args} shows graphical depictions of
example ARGs produced by four tools using substantially
different inference strategies.

% Indeed, a
% common statement in the literature is that modern tools such as
% \tsinfer~\citep{kelleher2019inferring} and \relate~\citep{speidel2019method}
% construct not an ARG but ``only'' a sequence of local trees
% \citep[e.g.][]{hejase2020summary, brandt2021evaluation, ignatieva2021kwarg}.
% While this statement is correct inasmuch as neither method attempts to infer
% the explicit recombination events that cause adjacent trees to differ, it is
% certainly not the case that they infer independent local trees. In fact, both
% methods take advantage of genome-wide mutation information to construct
% genealogies, and as a result nodes and edges persist across multiple local
% trees. Their results can therefore be thought of as an inheritance graph, and
% represented in the gARG format.

\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} Inference of sample-resolved ARGs for 11
\textit{Drosophila melanogaster} DNA sequences over a 2.4kb
region of the ADH locus~\citep{kreitman1983nucleotide}.
Results for four different methods:
(A) \kwarg; (B) \argweaver; (C) \tsinfer; and (D) \relate. See the
text for details of these methods.
Edge colours are based on the time of the edge's child node
(lighter: older; darker: younger).
Vertical and horizontal positions of graph nodes are arbitrary.
Line width and node colour are as described in Fig.~\ref{fig-simplification}.
Graphics on the bottom row show the genome positions, relative to the start of
the ADH gene, for each graph edge from the corresponding ARG. Edge intervals
are drawn as horizontal lines, stacked in time order (edges with youngest
children at the bottom); vertical dashed lines indicate breakpoints between
local trees.
}
\end{figure}

The first two methods explicitly infer recombination events.
\kwarg~\citep{ignatieva2021kwarg} is a parsimony based approach
which searches the space of plausible ARGs,
outputting minimal ones using heuristics.
\argweaver~\citep{rasmussen2014genome} on the other hand is
model-based, sampling from a discretised version of the
% could spell this out, but in reality everyone knows what the SMC
% is, and if they don't spelling it out won't help them.
SMC~\citep{mcvean2005approximating,marjoram2006fast}.
Note that both \kwarg\ and \argweaver\ produce many ARGs,
and those shown in Fig.~\ref{fig-inferred-args} are arbitrarily selected
examples.
While the second two methods
both produce a single best-guess estimate and
do not explicitly infer recombination events, they are based on
quite different principles.
\texttt{Tsinfer} works in a two-step process, first generating ancestral
haplotypes via heuristics and then inferring inheritance relationships
between them using the Li and Stephens model~\citep{li2003modeling}.
\texttt{Relate} works by [TODO: Ana can you fill in please?]
See Appendix~\ref{sec-survey-arg-infer} for more details on these
and other inference methods.

Inferred ARGs are based on the \citet{kreitman1983nucleotide} dataset,
a standard benchmark in the classical ARG
literature.
It consists of 43 biallelic SNPs spanning 2.4Kb of the \emph{D.~melanogaster} ADH
locus on chromosome 2L. Where required for inference purposes we assume a
mutation rate of $5.49\times10^{-9}$
and recombination rate of $8.4\times10^{-9}$ per site per generation
% NOTE: we should be citing the original sources here too.
\citep{adrion2020community,lauterbur2023expanding},
% Note: would have been simpler to use the stdposim estimate
% of 1.7M from Li et al 2006. (not that it matters)
and constant effective population size of $10^6$.
Full details of the software used and code to generate these
figures are available on GitHub (see Data Availability).

% Here we describe recombination features
Considering~Fig~\ref{fig-inferred-args}, we can see that
there is considerable variation in the number of recombination breakpoints
inferred by different methods, with e.g.~\argweaver\ suggesting
far more than the 7 required for this dataset under minimal parsimony
assumptions~\citep{song2003parsimonious}.
A sense of the amount of recombination in each ARG is provided by
the node colouring scheme, which shows the number of parents for each node.
In Fig.~\ref{fig-inferred-args}A,B, each recombination event
corresponds to a node with exactly 2 parents and one child.
As these methods explicitly infer a recombination event for each
breakpoint, the number of breakpoints equals the number of 2-parent
(brown) nodes.
In contrast, Fig.~\ref{fig-inferred-args}C,D do not have this
straightforward
relationship between the number of nodes with multiple parents
and breakpoints along the genome.
In both ARGs the number of breakpoints is smaller
than the number of multiple-parent ARG nodes, showing
that several multiple-parent nodes must share breakpoint positions.
There are also ARG nodes with multiple parents and multiple children,
where one or more recombinations have been pushed down onto a
more recent node.
% Several of these nodes have more than two
% % parents (in red), representing multiple recombinations inferred to have
% occurred in an unresolvable order.
% Nevertheless, the number of breakpoints in
% these two ARGs is smaller than the number of multiple-parent ARG nodes, showing
% that several multiple-parent nodes must share breakpoint positions.
As a consequence, it may be difficult to condense each transition between trees
in these ARGs into a set of SPR operations.
% The sharing
% of breakpoint locations between these nodes indicates that in \tsinfer\ and
% \relate\ ARGs it may be difficult to condense each transition between trees
% into a set of simple SPR moves.
% Overall, Fig.~\ref{fig-inferred-args} shows
% that \tsinfer\ and \relate\ ARGs contain a level of detail about recombination
% that lies somewhere between a sequence of unrelated local trees on one extreme
% and an ARG with precisely specified recombination events on the other.

% Here we move on to coalescence
Shading within nodes in Fig~\ref{fig-inferred-args} indicates the
fraction of the node's span over which it is coalescent
(Section~\ref{sec-locally-unary-edges}). For example, brown nodes
in Fig~\ref{fig-inferred-args}A,B are clear because there is
no local coalescence at these recombination nodes
(they are ``graph unary'', and so local coalescence is impossible).
The significant number of partially shaded nodes in
Fig~\ref{fig-inferred-args}A,B and C demonstates that
\kwarg, \argweaver\ and \tsinfer\ all contain locally unary nodes.
Another difference between methods highlighted in this figure
is the presence of polytomies, which only
\tsinfer\ creates. The most obvious example involves
involves nodes \noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F},
which happen happen to have identical sequences.
Because \kwarg, \argweaver, and \relate\ require
bifurcating trees by design, each picks an arbitrary order of branching
(hence Fig.~\ref{fig-inferred-args}A,B disagrees
with Fig.~\ref{fig-inferred-args}D in this order).
% JK: I don't quite get this, so readers probably won't either. Let's
% stick a few clear messages.
% More generally, the number of nodes in Fig.~\ref{fig-inferred-args}C
% that have more than two children hints at considerable uncertainty in
% the order of branching elsewhere in the ARG.

The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
the genome to which graph edges are shared between multiple trees.
All four methods infer nodes and edges that are shared between
multiple tree, to varying degrees.
For example, all of the methods infer
infer that \noderef{Af-f}, \noderef{Fr-f}, and \noderef{Wa-f}
form a clade along the entire sequence.
In particular, we can
see both \tsinfer\ and (to a lesser extent) \relate\ have edges
that span multiple tree boundaries, indicating that they are not inferring
a series of \emph{unrelated} local trees.
However, in comparison to \kwarg\ and \argweaver\ neither
method results in extensive node sharing in the
oldest time periods.
Overall, Fig.~\ref{fig-inferred-args} shows
that \tsinfer\ and \relate\ ARGs contain a level of detail
that lies somewhere between a sequence of unrelated local trees on one extreme
and an ARG with precisely specified recombination events on the other
(Fig~\ref{fig-inferred-args}A,B).
% JK: Let's not get into this.
% This primarily reflects the fact that they do not encode topology
% above the local MRCAs, also indicated by multiple roots in the ARG
% (nodes with zero parents, in blue), compared to the
% single ARG root in the \kwarg\ and \argweaver\ genealogies. In terms of inference,
% \tsinfer\ and \relate\ show a pragmatic representation: we do not expect
% genetic variation to be very informative about topological relationships
% between nodes that that are older than the local MRCAs.
